\documentclass[a4paper]{easychair}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{cmll}
\usepackage{mathpartir}
\usepackage{mathrsfs}
\usepackage{xcolor}

\def\newelims{1}
\input{../macros.tex}

\begin{document}

\title{Linear metatheory via linear algebra}

\author{
  Robert Atkey\inst{1}
  \and
  James Wood\inst{1}\thanks{James Wood is supported by an EPSRC Studentship.}
}

\institute{
  University of Strathclyde,
  Glasgow, United Kingdom \\
  \email{\{robert.atkey,james.wood.100\}@strath.ac.uk}
}

\titlerunning{Linear metatheory via linear algebra}
\authorrunning{Wood and Atkey}

\maketitle

\paragraph{Introduction.}

We introduce a simply typed calculus \name{} that allows the use of variables to
be constrained by usage annotations in the context which binds them.
\name{} is a generalisation of existing core type theories for sensitivity analysis
\cite{reed10distance}, dependency and confidentiality \cite{abadi99core},
linearity \cite{Barber1996}, and modal validity \cite{judgmental}.
It is related to quantitative type theory \cite{quantitative-type-theory}, and
various coeffect calculi \cite{PetricekOM14,BrunelGMZ14,GhicaS14}.

One of our insights is that because our usage annotations form a semiring, we
have just enough structure to talk about vectors and matrices in the metatheory.
We find useful some constructs of linear algebra, culminating in substitution
phrased as application of a linear map.

An earlier version of this work was presented at TyDe 2018
\cite{context-constrained}.
The syntax and semantics are formalised in Agda, with the code at
\url{https://github.com/laMudri/quantitative/}.

\paragraph{Syntax.}

Our syntax is that of a simply typed $\lambda$-calculus modified to let us
reason about how variables are used.
We assume a partially ordered semiring (posemiring) $(\mathscr R, \subres, 0,
+, 1, *)$ of usage annotations, with elements coloured in green for emphasis.
The types are base types ($\base_k$), functions ($\fun{}{}$), tensor products
($\tensorOne$, $\tensor{}{}$), with products ($\withTOne$, $\withT{}{}$), sums
($\sumTZero$, $\sumT{}{}$), and graded bangs ($\excl{\rho}{}$).
A context $\ctx{\Gamma}{R}$ is the combination of a typing context $\Gamma$ and
a usage context $\resctx R$.
We see a usage context as a vector over $\mathscr R$ generated by the basis made
of the variables it contains.

\vspace{-0.3cm}

\begin{mathpar}
  \rescomment \rho, \rescomment \pi \in \mathscr R \and
  A, B, C \Coloneqq \base_k \mid \fun{A}{B} \mid \tensorOne \mid \tensor{A}{B}
  \mid \withTOne \mid \withT{A}{B} \mid \sumTZero \mid \sumT{A}{B} \mid
  \excl{\rho}{A} \\
  \Gamma, \Delta \Coloneqq \cdot \mid \Gamma, x : A \and
  \resctx P, \resctx Q, \resctx R \Coloneqq
  \cdot \mid \resctx R, x^{\rescomment \rho} \and
  \ctx{\Gamma}{R} \Coloneqq
  \cdot \mid \ctx{\Gamma}{R}, \ctxvar{x}{A}{\rho}
\end{mathpar}

Tensor products are eliminated by pattern matching (each side bound with
annotation $1$), whereas with products are eliminated by projections.
The difference is correspondingly seen in the introduction rules, where the two
halves of a tensor product have separate usage, and the two halves of a with
product have shared usage (illustrated below).

The rule \TirName{$\otimes$-I} is the archetypal use of $+$.
The constraint $\subrctx{\resctx P + \resctx Q}{\resctx R}$ says that $\resctx
R$ must be at least as permissive as the accumulation of usages in $\resctx P$
and $\resctx Q$.
If the addition of the semiring is a join of the order (as in the modality
example below), these two types of product are equivalent.

\vspace{-0.3cm}

\begin{mathpar}
  \inferrule*[right=$\otimes$-I]
  {\ctx{\Gamma}{P} \vdash M : A
    \\ \ctx{\Gamma}{Q} \vdash N : B
    \\ \subrctx{\resctx P + \resctx Q}{\resctx R}
  }
  {\ctx{\Gamma}{R} \vdash \ten{M}{N} : \tensor{A}{B}}

  \and

  \inferrule*[right=$\&$-I]
  {\ctx{\Gamma}{R} \vdash M : A
    \\ \ctx{\Gamma}{R} \vdash N : B
  }
  {\ctx{\Gamma}{R} \vdash \wth{M}{N} : \withT{A}{B}}
\end{mathpar}

With the graded bang, we see use of $*$ from the annotation posemiring.
We read $\rescomment \rho * \rescomment \pi$ as applying the action $\rescomment
\rho$ to $\rescomment \pi$.
Introduction can be seen as scaling usage.
Elimination is by pattern matching, where we bind a new variable with whatever
usage annotation the type gave us.

\vspace{-0.3cm}

\begin{mathpar}
  \inferrule*[right=$\excl{\rho}{}$-I]
  {\ctx{\Gamma}{P} \vdash M : A
    \\ \subrctx{\rescomment \rho * \resctx P}{\resctx R}
  }
  {\ctx{\Gamma}{R} \vdash \bang{M} : \excl{\rho}{A}}
  \and
  \inferrule*[right=$\excl{\rho}{}$-E]
  {\ctx{\Gamma}{P} \vdash M : \excl{\rho}{A}
    \\ \ctx{\Gamma}{Q}, \ctxvar{x}{A}{\rho} \vdash N : B
    \\ \subrctx{\resctx P + \resctx Q}{\resctx R}
  }
  {\ctx{\Gamma}{R} \vdash \bm{B}{M}{x}{N} : B}
\end{mathpar}

The \TirName{var} rule (not pictured) at $x$ can only be used in a usage context
$\resctx R$ when $x$ can be used plainly, and all other variables can be
discarded.
This can be succinctly stated as the constraint $x^1 \subres \resctx R$,
where $x^1$ is the $x$th basis vector.

\vspace{-0.1cm}

\paragraph{Substitution.}

We have two admissible rules leading up to the substitution lemma --- subusaging
(\TirName{subuse}) and weakening (\TirName{weak}) --- stated below.
In the language of linear algebra, weakening is embedding into a space of higher
dimension.

Let $|-|$ denote the length of a context.
Usage contexts are taken to be row vectors.
A \emph{substitution} $\sigma$ from $\ctx{\Gamma}{P}$ to $\ctx{\Delta}{Q}$
comprises a $|\resctx Q| \times |\resctx P|$ matrix $\mat \Sigma$ such that
$\resctx Q \mat \Sigma \subres \resctx P$, and for each $(x:A) \in \Delta$, a
term $M_x$ such that $\Gamma^{x^1 \mat \Sigma} \vdash M_x : A$.
Then, the simultaneous substitution lemma is proven via the linearity of
vector-matrix multiplication.

\vspace{-0.6cm}

\begin{mathpar}
  \inferrule*[lab=subuse]{
    \ctx{\Gamma}{P} \vdash M : A
    \\ \resctx P \subres \resctx Q
  }
  {
    \ctx{\Gamma}{Q} \vdash M : A
  }

  \and

  \inferrule*[lab=weak]{
    \ctx{\Gamma}{P} \vdash M : A
  }
  {
    \ctx{\Gamma}{P}, \Delta^{\vct 0} \vdash M : A
  }

  \and

  \inferrule*[right=subst]{
    \ctx{\Delta}{Q} \vdash N : A
    \\\\ \sigma : \ctx{\Gamma}{P} \Rightarrow \ctx{\Delta}{Q}
  }
  {
    \ctx{\Gamma}{P} \vdash N[\sigma] : A
  }
\end{mathpar}

\vspace{-0.2cm}

The identity substitution, where each variable $x$ is substituted by the term
$x$, is witnessed by the identity matrix.
We expect composition to be witnessed by matrix multiplication.

\vspace{-0.1cm}

\paragraph{Specialisations.}
\label{sec:specialisations}

To demonstrate the applicability of \name{}, and give examples of usage
posemirings, we have shown that certain instances are translatable to DILL
\cite{Barber1996} and the modal type theory of Pfenning and Davies
\cite{judgmental}.
Future work is to give an equational theory for \name{}, and show that these
translations form an isomorphism.

\vspace{-0.1cm}

\paragraph{Semantics.}
We also have a semantics that captures the intensional properties of programs
via families of Kripke indexed relations that refine a simple set-theoretic
semantics.
This allows us to reconstruct the semantic properties of calculi in prior work
for sensitivity analysis \cite{reed10distance}, and dependency and
confidentiality \cite{abadi99core}, as well as a new calculus for monotonicity
analysis.

\bibliographystyle{alpha}
\bibliography{../paper/quantitative}

\end{document}